---
title: "Simulating contact tracing data"
author: "Thibaut Jombart"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: spacelab
    highlight: pygments
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, fig.width = 8, fig.height = 6, dpi = 70)
```



# Data simulation

## Simulated distributions

### Incubation period

```{r }

library(distcrete)
library(epitrix)
library(tidyverse)

incub_mu <- 9.4
incub_sd <- 7.4

params <- gamma_mucv2shapescale(incub_mu,
                                      incub_sd / incub_mu)

incubation <- distcrete("gamma",
                        interval = 1,
                        shape = params$shape,
                        scale = params$scale)

```

Example:

```{r }

incubation$r(1000) %>%
  qplot() +
  theme_bw() +
  labs(x = "Incubation time",
       y = "Frequency")

```



### Duration of exposures

Duration is drawn from a distribution $~ 1 + Exp(d - 1)$ where $d$ is the mean
duration of exposure (must be at least 1).

```{r }

duration_r <- function(n, mean = 1) {
  if (mean < 1) {
    msg <- "mean duration of exposure must be >=1 day"
    stop(msg)
  }
  
  1L + rpois(n, mean - 1)
}

```

Example:

```{r }

duration_r(1000, mean = 2.5) %>%
  qplot() +
  theme_bw() +
  labs(x = "Duration of exposure",
       y = "Frequency")

```



### Dates of exposures

Date of exposure are drawn from a Uniform distribution on a given time-frame:

```{r }

date_exposure_r <- function(n, min = 0L, max = 30L, mean_duration = 1) {
  base <- seq(from = min, to = max, by = 1L)
  L <- length(base)
  date_start <- base[sample.int(L, size = n, replace = TRUE)]
  durations <- duration_r(n, mean_duration)
  data.frame(
      id = as.integer(1:n),
      exposure_start = as.integer(date_start),
      exposure_end = as.integer(date_start + durations - 1),
      exposure_duration = as.integer(durations)
  )
}

```

Example:

```{r }

toy_data <- date_exposure_r(10, mean_duration = 1.5)
toy_data

toy_data %>%
  mutate(id = factor(id)) %>%
  ggplot(aes(x = exposure_start, y = id, color = id)) +
  geom_segment(aes(xend = exposure_end, yend = id)) +
  geom_point() +
  geom_point(aes(x = exposure_end)) +
  theme_bw() +
  guides(color = FALSE) +
  labs(x = "Exposure windows", y = "Contact ID")

```



### Infection of contacts

The infection of contacts is determined as follows:

1. determine the probability of infection given the rate of infectious contacts
2. determine if there is an infection using a Binomial (Bernoulli) distribution
3. determine the day of the infection from the window of exposure
4. determine the day of symptom onset from the incubation time distribution

```{r }


sample_ <- function(x, n = length(x), replace = FALSE) {
  x[sample.int(length(x), n, replace = replace)]
}

pick_exposure <- function(start, end) {
  vapply(seq_along(start),
         function(i) sample_(start[i]:end[i], 1L),
         1L)
}

#' @param x contact data as returned by `date_exposure_r`
#'
#' @param rate_infection a positive rate of infection per day of contact
#'
#' @param incubation_time a function simulating incubation times whose only
#'   argument is `n`, the number of simulated incubation times
#' 
infect_contacts <- function(x,
                            rate_infection = 0,
                            incubation_time = incubation$r) {

  out <- mutate(
      x,
      p_infection = 1 - exp(-rate_infection * exposure_duration),
      rnd = runif(n()),
      case = if_else(rnd < p_infection, "case", "non_case"),
      case = factor(case, levels = c("case", "non_case")),
      date_infection = if_else(
          case == "case",
          pick_exposure(exposure_start, exposure_end),
          NA_integer_),
      date_onset = if_else(
          case == "case",
          as.integer(date_infection + incubation_time()),
          NA_integer_))

  select(out, -rnd)
}

```

Example:

```{r }

toy_data <- infect_contacts(toy_data, rate_infection = 0.7)
toy_data

toy_data %>%
  pivot_longer(cols = date_infection:date_onset,
               names_to = "event",
               values_to = "event_date") %>%
  mutate(event = sub("date_", "", event)) %>% 
  mutate(id = factor(id)) %>%
  ggplot(aes(x = exposure_start, y = id, color = case)) +
  geom_segment(aes(xend = exposure_end, yend = id)) +
  geom_point() +
  geom_point(aes(x = exposure_end)) +
  geom_point(aes(x = event_date, shape = event), size = 4) +
  scale_shape_manual(values = c(infection = 4, onset = 1)) +
  theme_bw() +
  labs(x = "Time", y = "Contact ID")

```



## Contact simulator

The contact data simulator assembles the various components of the functions
above, and outputs S3 objects with class `contact_data`.

```{r }

#' Simulate contact tracing data
#' 
#' @param n the number of contacts to simulate
#' 
#' @param rate_infection a positive rate of infection per day of contact
#'
#' @param
#'

simulate_contacts <- function(n,
                              incubation,
                              rate_infection = 0,
                              exposure_mean_duration = 1,
                              exposure_min_date = 0,
                              exposure_max_date = 30
                              ) {

  out <- date_exposure_r(n = n,
                         min = exposure_min_date,
                         max = exposure_max_date,
                         mean_duration = exposure_mean_duration)
  
  out <- infect_contacts(out, rate_infection = rate_infection)
  out <- tibble(out)
  class(out) <- c("contact_data", class(out))
  out
}


plot.contact_data <- function(x) {
  
  pivot_longer(x,
               cols = date_infection:date_onset,
               names_to = "event",
               values_to = "event_date") %>%
  mutate(event = sub("date_", "", event)) %>% 
  mutate(id = factor(id)) %>%
  ggplot(aes(x = exposure_start, y = id, color = case)) +
  geom_segment(aes(xend = exposure_end, yend = id)) +
  geom_point() +
  geom_point(aes(x = exposure_end)) +
  geom_point(aes(x = event_date, shape = event), size = 4) +
  scale_shape_manual(values = c(infection = 4, onset = 1)) +
  theme_bw() +
  labs(x = "Time", y = "Contact ID")
}

```

Example:

```{r }

x <- simulate_contacts(50,
                       incubation = incubation$r,
                       rate_infection = .2,
                       exposure_mean_duration = 5)
x
plot(x)

```





# Contact followup simulation

## Adding followup info to contacts

This function adds to variables used for the followup of contact to
`contact_data` objects; these variables include when followup must start and end
(both dates inclusive), when contacts were last seen, if cases have been
detected, and when they were..

```{r }

add_followup <- function(x, followup_duration) {
  
  ## add needed variables on followup history

  ## * followup_start and _end: time window for active followup (both dates
  ##   included)
  ## * last_seen: needed to define 'detected' cases
  ## * detected: needed to ignore already detected cases
  ## * detected_at: needed to calculate delays to detection
  
  mutate(x,
         followup_start = as.integer(exposure_end + 1L),
         followup_end = as.integer(followup_start + followup_duration - 1L),
         last_seen = rep(-1L, n()),
         detected = rep(FALSE, n()),
         detected_at = rep(NA_integer_, n()))
}

x <- x %>%
  add_followup(21)
x

```


## Finding active contacts

Active contacts are contacts which, at time *t*:

* have not been detected yet
* are within the followup window

```{r }

active_contacts <- function(x, t, filter = FALSE) {
  if (!"detected" %in% names(x)) {
    x <- mutate(x, detected = rep(FALSE, n()))
  }
 
  out <- mutate(x,
                active =
                  t >= followup_start &
                  t <= followup_end &
                  !detected)
  if (filter) {
    out <- filter(out, active)
    out <- select(out, -active)
  }
  out
}


```

Example:

```{r }

x %>%
  active_contacts(5)

x %>%
  active_contacts(10)

x %>%
  active_contacts(10, filter = TRUE) %>%
  plot()

```



## Contact followup implementation {.tabset .tabset-fade .tabset-pills}

In this section we devise functions which will process a `contact_data` object
and simulate follow-up over time using a given contact followup
strategy. Strategies themselves are functions with a pre-defined set of
arguments, taking contact tracing data in and outputting a IDs of contacts to
follow. The function `check_strategy` will be used to check that a `strategy`
has the correct set of arguments.

### Check strategy

```{r }

check_strategy <- function(f) {
  if (!is.function(f)) {
    msg <- "`f` is not a function"
    stop(msg)
  }
  
  ## check signature
  arguments <- as.list(args(f))
  if (!identical(
           names(arguments)[1:3],
           c("x", "n", "t"))
      ) {
    msg <- "arguments must be: `x`, `n` and `t`"
    stop(msg)
  }
}

```


### Followup at a given time

This function achieves followup at a given time *t*, taking into account past
followup history if it is contained in the data.

```{r }

#' Followup at time t
#'
#' This function will perform contact followup at a given time `t` given a
#' user-provided strategy.
#' 
#' @param x a `contact_data` object, potentially containing previous followup
#'   history
#'
#' @param n_followed the number of contacts followed at time `t`
#'
#' @param t current time of followup
#'
#' @param strategy a function taking as input a `contact_data` object `x`, the
#'   number of contacts to follow `n`, and the current time `t`
#'
#' @param followup_only a `logical` indicating if only followup information
#'   should be returned; if `FALSE` (default), the original data with added
#'   followup information is returned

followup_t <- function(x, n_followed, t,
                       strategy = random_strategy,
                       followup_only = FALSE) {

  ## perform checks
  check_strategy(strategy)
  
  ## add variables as needed
  t <- as.integer(t)

  ## add needed variables on followup history if they are missing
  ## * last_seen: needed to define 'detected' cases
  ## * detected: needed to ignore already detected cases
  if (!"last_seen" %in% names(x)) {
    x <- mutate(x, last_seen = rep(-1L, n()))
  }
  if (!"detected" %in% names(x)) {
    x <- mutate(x, rep(FALSE, n()))
  }
  if (!"detected_at" %in% names(x)) {
    x <- mutate(x, detected_at = rep(NA_integer_, n()))
  }

  ## check active contacts
  out <- active_contacts(x, t, filter = FALSE)

  ## make sure we don't follow more contacts than there are active ones
  n_active <- sum(out$active)
  n_followed <- min(n_followed, n_active)
  
  ## find which ones to follow
  id_followed <- strategy(out, n_followed, t)

  
  ## add followup info
  out <- mutate(
      out,
      followup = if_else(id %in% id_followed,
                         "seen",
                         "missed"),
      followup = if_else(active, followup, "inactive"),
      followup = if_else(followup == "seen" &
                         t >= date_onset &
                         last_seen < date_onset &
                         case == "case",
                         "detected",
                         followup),
      detected = if_else(followup == "detected",
                         TRUE,
                         detected),
      detected_at = if_else(followup == "detected",
                         t,
                         detected_at),
      last_seen = if_else(followup %in% c("detected", "seen"),
                          t,
                          last_seen)
      )
  out <- rename(out, !!paste("followup", t, sep = "_") := followup)
  out <- select(out, -active)
  if (followup_only) {
    out <- select(out, !!paste("followup", t, sep = "_"))
  }
  out
}

```



### Followup over successive time windows

This is the main function to implement followup. For convenience, we also build
a function which will achieve followup for all days in a given time window:

```{r }

followup <- function(x, coverage, strategy = random_strategy) {
  if (!inherits(x, "contact_data")) {
    msg <- "x is not a `contact_data` object"
  }
  check_strategy(strategy)
  
  days <- seq(from = min(x$followup_start), to = max(x$followup_end), by = 1L)
  n_followed <- round(coverage * nrow(x))

  previous_followup <- x
  
  for (i in seq_along(days)) {
    out <- followup_t(previous_followup,
                      t = days[i],
                      n_followed = n_followed,
                      strategy = strategy)
    previous_followup <- out
  }

  ## set detected in non_cases to NA
  out <- mutate(out, detected = if_else(case == "case", detected, as.logical(NA)))
  
  class(out) <- c("followup", class(out))
  out
}

```


### S3 methods

We also add a function to plot `followup` objects:

```{r }

plot.followup <- function(x, show_inactive = TRUE) {
  
  ## color scale
  scale_fill_followup <- scale_fill_manual(
      values = c(inactive = "#C4B5B2",
                 detected = "#B74967",
                 seen = "#49A0D5",
                 missed = "#ECBF74")
  )

  ## reshape data into long format
  x_long <- select(x,
                   id,
                   contains("followup"),
                   -followup_start,
                   -followup_end)
  x_long <- pivot_longer(x_long, -1, names_to = "time", values_to = "followup")
  x_long <- mutate(x_long,
                   time = gsub("followup_", "", time),
                   time = as.integer(time),
                   followup = factor(
                       followup,
                       levels = c("inactive", "seen", "missed", "detected")))

  if (!show_inactive) {
    x_long <-  filter(x_long, followup != "inactive")
  }
  
  ## make plot
  x_counts <- count(x_long, time, followup)
  ggplot(x_counts, aes(x = time, y = n)) +
  geom_col(aes(fill = followup)) +
  theme_bw() +
  scale_fill_followup +
  labs(x = "time",
       y = "number of contacts",
       title = "Contact followup")
}

```

A summary method will provide information on:

* the breakdown of cases detected / not detected, and associated proportion and CI
* data on delay to detection (onset to detection)
* summary on delay to detection

```{r }

summary.followup <- function(object) {
  x <- object
  cases <- filter(x,
                  case == "case")

  ## breakdown of cases
  n_cases <- nrow(cases)
  n_detected <- sum(cases$detected)
  n_non_detected <- sum(!cases$detected)

  ## delay to detection
  delay_detection <- as.integer(cases$detected_at - cases$date_onset)
  delay_detection <- delay_detection[!is.na(delay_detection)]

  list(
      n_contacts = nrow(x),
      n_cases = n_cases,
      n_detected = n_detected,
      n_non_detected = n_non_detected,
      p_detected = n_detected / n_cases,
      p_detected_ci = prop.test(n_detected, n_cases)["conf.int"]$conf.int,
      delay_detection = delay_detection,
      delay_detection_smry = summary(delay_detection)
  )
}

```



# Testing followup strategies {.tabset .tabset-pills}

## Random strategy

### Strategy implementation

```{r }

#' Random contact followup
#'
#' @param x a `contact_data` with information on active contacts as returned by
#'   `active_contacts`
#'
#' @param n the number of contacts to be followed
#'
#' @param t the current time; not used, but present for compatibility with other methods
#'
#' @return a vector of IDs of contacts to be followed
#' 
random_strategy <- function(x, n, t) {
  ## strategy:
  ## 1. determine which contacts are still active
  ## 2. sample IDs randomly from active IDs

  x <- filter(x, active)
  x <- sample_n(x, n)
  pull(x, id)
  
}

```





### Example

Here we run an example based on 10,000 contacts to follow.

```{r }

res_random <- simulate_contacts(
    10000,
    incubation = incubation$r,
    rate_infection = .2,
    exposure_mean_duration = 3) %>%
  add_followup(incubation$q(.95)) %>%
  followup(.01)

plot(res_random, FALSE)


## proportion of cases detected
summary(res_random)$p_detected
summary(res_random)$p_detected_ci

## delay to detection
data.frame(x = summary(res_random)$delay_detection) %>%
  ggplot(aes(x = x)) +
  geom_density(color = "#ECBF74", fill = "#ECBF74", alpha = .6) +
  theme_bw() +
  labs(x = "Days from onset to detection",
       y = "Frequency",
       title = "Distribution of delay to detection")

```




## New first strategy

In this strategy, contacts who reported the most recent exposure are prioritised.

### Strategy implementation

```{r }

#' new first contact followup
#'
#' @param x a `contact_data` with information on active contacts as returned by
#'   `active_contacts`
#'
#' @param n the number of contacts to be followed
#'
#' @param t the current time; not used, but present for compatibility with other methods
#'
#' @return a vector of IDs of contacts to be followed
#' 
new_first_strategy <- function(x, n, t) {
  ## strategy:
  ## 1. keep only active contacts
  ## 2. calculate delay between last exposure and current time
  ## 3. sort them by decreasing delay
  ## 4. keep the first n ids
 
  x <- filter(x, active)
  x <- mutate(x, delay = as.integer(t - followup_start))
  x <- arrange(x, delay)
  x <- slice(x, seq_len(n))
  pull(x, id)
  
}

```


### Example

Here we run an example based on 10,000 contacts to follow.

```{r }

res_new_first <- simulate_contacts(
    10000,
    incubation = incubation$r,
    rate_infection = .2,
    exposure_mean_duration = 3) %>%
  add_followup(incubation$q(.95)) %>%
  followup(.01, strategy = new_first_strategy)

plot(res_new_first, FALSE)


## proportion of cases detected
summary(res_new_first)$p_detected
summary(res_new_first)$p_detected_ci

## delay to detection
data.frame(x = summary(res_new_first)$delay_detection) %>%
  ggplot(aes(x = x)) +
  geom_density(color = "#ECBF74", fill = "#ECBF74", alpha = .6) +
  theme_bw() +
  labs(x = "Days from onset to detection",
       y = "Frequency",
       title = "Distribution of delay to detection")

```








## No drop strategy

In this strategy, contacts who were seen the most recently are prioritised so
that continuous followup of a few contacts is ensured. When no contacts have
been seen, we revert to the "new first" strategy.

### Strategy implementation

```{r }

#' No drop contact followup
#'
#' @param x a `contact_data` with information on active contacts as returned by
#'   `active_contacts`
#'
#' @param n the number of contacts to be followed
#'
#' @param t the current time; not used, but present for compatibility with other methods
#'
#' @return a vector of IDs of contacts to be followed
#' 
no_drop_strategy <- function(x, n, t) {
  ## strategy:
  ## 1. keep only active contacts
  ## 2. calculate delay between last exposure and current time
  ## 3. sort them by decreasing delay
  ## 4. keep the first n ids
 
  if (!"last_seen" %in% names(x)) {
    return(new_first_strategy(x, n, t))
  }
  
  x <- filter(x, active)
  x <- mutate(x, delay = as.integer(t - last_seen))
  x <- arrange(x, delay)
  x <- slice(x, seq_len(n))
  pull(x, id)
  
}

```


### Example

Here we run an example based on 10,000 contacts to follow.

```{r }

res_no_drop <- simulate_contacts(
    10000,
    incubation = incubation$r,
    rate_infection = .2,
    exposure_mean_duration = 3) %>%
  add_followup(incubation$q(.95)) %>%
  followup(.01, strategy = no_drop_strategy)

plot(res_no_drop, FALSE)


## proportion of cases detected
summary(res_no_drop)$p_detected
summary(res_no_drop)$p_detected_ci

## delay to detection
data.frame(x = summary(res_no_drop)$delay_detection) %>%
  ggplot(aes(x = x)) +
  geom_density(color = "#ECBF74", fill = "#ECBF74", alpha = .6) +
  theme_bw() +
  labs(x = "Days from onset to detection",
       y = "Frequency",
       title = "Distribution of delay to detection")

```
